Ownership types in theory and practice (in Rust)

### 2023-01-22 paper/paper.tex
More progress on slides pt 2
finished transferring ownership slides
worked on references borrowing slide

% TODO: brief overview of sections somewhere (maybe end of introduction)

@@

One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
However, as will be discussed further in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
This allows for expressing more programming idioms.

@@

\paragraph{Heap-based variables}

@@

Immutable references can be seen as read-only references and are acquired using the \emph{\&} operator.
The value that immutable references refer to may not be modified.
Otherwise the program will not compile.

@@

Rust also supports mutable references, which are created with the \emph{\&mut} keyword and allow the borrowed value to be modified.
For obvious reasons, creating a mutable reference is only possible if the borrowed value is also mutable.

@@

This restriction exists to identify and prevent \emph{data races}, which can cause undefined behavior at compile time.
A data race occurs when two or more pointers access the same data simultaneously, at least one of the points is being used to write the data, and there exists no mechanism to synchronize access to the data \cite{rust-book}.

@@

% TODO: explain how dangling references are prevented

% explain this example:
% \begin{lstlisting}
% fn dangle() -> &String {
%   let s = String::from("hello world");
%   &s
% }
% \end{lstlisting}

>>>>

% TODO: brief overview of sections somewhere (maybe end of introduction)
% TODO: replace \emph with \verb

@@

One can argue that Rust's ownership system is an extended version of the ownership model \emph{owners-as-modifiers}.
The Rust and owners-as-modifiers approaches differentiate themselves from the owners-as-dominators model, since they both allow read-only references to any object.
Furthermore, as will be discussed in \ref{sec:rust-references}, Rust allows both read-only and mutable references in a controlled manner.
This allows for expressing more programming idioms.

@@

\paragraph{Heap-based values}

@@

Immutable references can be seen as read-only references and are acquired using the \verb|&| operator.
The value that immutable references refer to may not be modified.
Otherwise the program will not compile.

@@

Rust also supports mutable references, which are created with the \verb|&mut| keyword and allow the borrowed value to be modified.
For obvious reasons, creating a mutable reference is only possible if the borrowed value is also mutable.

@@

This restriction exists to identify and prevent \emph{data races} at compile time, which can cause undefined behavior.
A data race occurs when two or more pointers access the same data simultaneously, at least one of the pointers is being used to write the data, and there exists no mechanism to synchronize access to the data \cite{rust-book}.

@@

### 2023-01-22 slides/slides.tex
worked on references borrowing slide

\begin{frame}{References and borrowing}
\end{frame}

@@

@@

>>>>

@@

\begin{frame}{Borrowing}
\begin{itemize}
  \item \emph{reference}: an address that points to a value that is owned by another variable
  \item unlike \emph{pointers}, a reference is guaranteed to point to a valid value
  \item creating a reference in Rust is called \emph{borrowing}
  \item a reference does not own the value it points to $\rightarrow$ only reference (not the borrowed value) is dropped at the end of the reference's scope
  \item used for performing operations on values without taking ownership
  % \item the scope of a reference starts where it is introduced and extends until the last time it is used
  \item Rust supports immutable and mutable references
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing: immutable references}
\begin{itemize}
  \item read-only/shared references $\rightarrow$ compiler disallows modifying borrowed value
  \item acquired using the \verb|&| operator
  \item unlimited amount of immutable references to same value are allowed (if no mutable references)

  \begin{lstlisting}[language=Rust]
  let s = String::from("hello");
  let r1 = &s; // 1. reference
  let r2 = &s; // 2. reference
  println!("r1: {}, r2: {}", *r1, *r2);
  \end{lstlisting}
\end{itemize}
\end{frame}

% TODO: combine this and next slide
\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  \item borrowed value may be modified
  \item acquired using the \verb|&mut| keyword
  \item borrowed value must be mutable with \verb|mut| (variables are immutable by default)

  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  s.push_str(" world");
  String::push_str(&mut s, " world"); // equivalent
  \end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing: mutable references}
\begin{itemize}
  \item \emph{data races}: multiple pointers access the same data simultaneously and at least one is writing to the data
  \item goal: prevent \emph{data races} at compile time
    % TODO: latex beamer make stand out/highlight
  \item lifetime of mutable reference may not overlap with lifetime of any other (immutable or mutable) reference
  \item only one mutable reference to a value allowed at any given time
\end{itemize}
\end{frame}

@@

% TODO: will it compile: create multiple mutable references, but only use one
% However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.

%   \begin{lstlisting}[language=Rust]
%   let mut s = String::from("hello");

%   let r1 = &s; // allowed
%   let r2 = &s; // allowed
%   println!("{} and {}", r1, r2);

%   let r3 = &mut s; // allowed
%   r3.push_str(" world");
%   println!("{}", r3);
%   \end{lstlisting}

% Since \emph{r1} and \emph{r2} are not used after line 5, that is where their scopes end.
% The scope of the mutable reference \emph{r3} only starts in line 7.
% Therefore, the scopes of the mutable reference and the two immutable references do not overlap, and the program compiles.
